\chapter{Introduction}

The advent of intelligent transportation systems (ITS) hinges on robust and responsive Vehicle-to-Everything (V2X) communication, which facilitates the exchange of safety, traffic efficiency, and infotainment data between vehicles (V2V), infrastructure (V2I), pedestrians (V2P), and networks (V2N). Among these, safety-critical applications—such as cooperative collision avoidance, intersection movement assist, and platooning—impose Ultra-Reliable Low-Latency Communication (URLLC) constraints. These services demand end-to-end latency guarantees on the order of milliseconds (typically 3-10 ms for the most critical alerts) and packet delivery reliability exceeding 99\% in high-density scenarios \cite{3gpp_ts_22_261_r18}. While dedicated standards like IEEE 802.11p (DSRC) and 3GPP Cellular-V2X (C-V2X) were explicitly designed for these low-latency, high-mobility environments, the pervasive deployment and low cost of commodity IEEE 802.11 (Wi-Fi) hardware present a compelling motivation to investigate its potential for deterministic V2X operation, especially in scenarios involving coexistence or cost-sensitive deployments \cite{peha2023v2x_shared_unlicensed}.

However, conventional Wi-Fi, built upon the Distributed Coordination Function (DCF) with Carrier Sense Multiple Access with Collision Avoidance (CSMA/CA), is fundamentally probabilistic and non-deterministic. Its random backoff mechanism, while providing statistical fairness, introduces unbounded channel access delay and jitter under increasing network load or contention. This inherent variability violates the strict worst-case latency requirements of safety-critical V2X services. The Enhanced Distributed Channel Access (EDCA) mechanism, introduced in IEEE 802.11e \cite{ieee80211e}, mitigates this by defining Access Categories (ACs) with distinct arbitration inter-frame space (AIFS), contention window (CW), and transmission opportunity (TXOP) parameters. While EDCA provides statistical prioritization, it remains a contention-based protocol. Under congestion, even the highest-priority traffic can experience unbounded delays due to repeated collisions and exponential backoff, failing to provide deterministic worst-case bounds.

In parallel, the field of wired networking has addressed similar challenges through Time-Sensitive Networking (TSN), a set of IEEE 802.1 standards that bring determinism to Ethernet. Key mechanisms like the Time-Aware Shaper (TAS) in IEEE 802.1Qbv \cite{ieee8021qbv} use scheduled, periodic gate-control lists (GCLs) to create protected time windows for critical traffic, effectively eliminating contention. Recent research efforts, often termed Wireless TSN, seek to extend these principles to Wi-Fi \cite{seliem2023wireless}. However, many of these proposals depend on advanced PHY/MAC features such as Orthogonal Frequency-Division Multiple Access (OFDMA), Target Wake Time (TWT), or specific firmware-level schedulers that are absent in the vast installed base of legacy 802.11a/b/g/n/ac devices. This reliance limits their immediate applicability and increases deployment complexity and cost \cite{schneider2022proposal}.

This work introduces \textbf{Deterministic QoS over Wi-Fi (DQ-WiFi)}, a novel MAC-layer framework engineered to deliver provable, bounded latency on legacy IEEE 802.11 devices. DQ-WiFi achieves this by architecturally decoupling deterministic service from hardware-dependent features, instead leveraging coordinated behavioral control across the network. The core innovation is the coupling of two complementary mechanisms: (i) a proactive, periodic \textbf{bandwidth reservation} scheme implemented via coordinated \textbf{LISTENING} windows, which temporarily silence all non-critical traffic to create collision-free transmission opportunities, and (ii) a reactive \textbf{BLOCKED} interval mechanism that immediately quenches network-wide contention upon the detection of an emergency event, preventing collision cascades.

Orchestrating these mechanisms is a lightweight, beacon-driven Finite State Machine (FSM) that runs identically on all nodes—Access Points (APs) and stations (STAs). This FSM enforces network-wide, synchronized transitions between three fundamental states (\textbf{ACTIVE}, \textbf{LISTENING}, and \textbf{BLOCKED}) based on timing parameters broadcast by a central Deterministic QoS Controller (DQC), typically the AP. This approach provides temporal partitioning without the need for complex per-flow reservations, admission control, or changes to the underlying hardware. The result is a system that guarantees collision-free service windows for time-critical traffic and provides deterministic latency bounds for aperiodic best-effort flows. DQ-WiFi is fully compatible with the extensive legacy ecosystem of 802.11a/b/g/n/ac devices, while its design is forward-compatible, allowing its temporal windows to be naturally mapped onto dedicated Resource Units (RUs) in future OFDMA-based Wi-Fi 6/7 implementations.

\textbf{Contributions.}
This thesis makes the following primary contributions:
\begin{enumerate}
\item The design of a \textbf{beacon-parameterized Finite State Machine (FSM)} that enforces cyclic \textbf{ACTIVE/LISTENING} operation for periodic bandwidth reservation and integrates a reactive emergency \textbf{BLOCKED} state for immediate contention containment.
\item A \textbf{formal worst-case delay analysis} that decomposes the end-to-end latency for both emergency and regular critical frames, providing analytical bounds based on the FSM's timing parameters and standard MAC-layer delays.
\item A \textbf{full-stack implementation} within the OMNeT++/INET simulation framework, including custom traffic generators (UdpCritApp, UdpCollisionApp), a DSCP-based QoS classifier, and the EdcaFsmController module, all seamlessly integrated into the standard HCF/EDCA architecture.
\item An \textbf{extensive evaluation} demonstrating that DQ-WiFi can achieve sub-millisecond latency for emergency alerts and provably bounded latency for aperiodic control traffic, significantly outperforming standard EDCA in deterministic performance while maintaining high channel utilization.
\end{enumerate}

The remainder of this thesis is structured as follows: Chapter 2 provides the necessary background on Wi-Fi operation, QoS mechanisms, and real-time systems theory. Chapter 3 surveys related work in deterministic networking. Chapter 4 formalizes the DQ-WiFi system model and delay analysis. Chapter 5 details the implementation. Chapter 6 presents and discusses the preliminary evaluation results. Finally, Chapter 7 outlines the work plan and concludes.