\chapter{Background}
\label{chap:background}

This chapter provides the technical foundation for the development and analysis of deterministic communication mechanisms over IEEE~802.11 networks, especially in Vehicle-to-Everything (V2X) contexts. While traditional Wi-Fi was not designed with timing guarantees in mind, recent evolutions in the standard, along with cross-domain contributions from real-time systems and TSN, have opened paths to bounded-latency wireless communication. 

We begin by surveying key features of modern IEEE~802.11 standards, particularly the PHY and MAC-layer enhancements introduced in 802.11ac/ax, including OFDMA, MU-MIMO, spatial reuse, and power-saving mechanisms such as Target Wake Time (TWT). These features shape the behavior of contention and scheduling in current Wi-Fi networks.

Next, we examine the Quality-of-Service (QoS) toolbox embedded in IEEE~802.11e and later amendments. Mechanisms like Enhanced Distributed Channel Access (EDCA), Hybrid Coordination Function (HCF), and TXOP-based reservations are discussed, highlighting their trade-offs in terms of efficiency, fairness, and latency isolation.

We then step into the domain of real-time scheduling theory, which forms the analytical basis for evaluating latency guarantees. We describe classical server-based models—periodic, sporadic, and aperiodic—as well as fixed-priority and EDF scheduling, focusing on their utilization bounds and schedulability conditions.

Building on these foundations, we present the Time-Sensitive Networking (TSN) framework as defined by the IEEE~802.1 working group. Standards such as 802.1AS (clock synchronization), 802.1Qbv (Time-Aware Shaping), and others are explored, showing how they enable deterministic delivery over Ethernet and, by extension, inspire mechanisms for wireless domains.

We then contextualize these mechanisms in the safety-critical requirements of V2X communications. URLLC targets for latency, reliability, and packet loss are summarized, along with benchmarks from both IEEE~802.11p (DSRC) and 3GPP-based C-V2X. The baseline operation of IEEE~802.11p is also presented as a reference point, emphasizing its low-latency design principles and limitations.

Finally, we describe the simulation environment used in this dissertation, based on the OMNeT++/INET framework, including Wi-Fi models, mobility constraints, and PHY/MAC abstraction layers.

The chapter concludes with a focused analysis of bounded latency in IEEE~802.11 networks.

% % 1 ─────────── Wi-Fi foundations
% \section{IEEE\,802.11 Fundamentals}
%   \subsection{From 802.11ac to 802.11ax: OFDMA, MU-MIMO, Spatial Reuse}
%   \subsection{Target Wake Time (TWT) and Beacon-driven Scheduling}

% % 2 ─────────── Built-in QoS
% \section{QoS Mechanisms in IEEE\,802.11}
%   \subsection{Enhanced Distributed Channel Access (EDCA)}
%   \subsection{Hybrid Coordination Function (HCF) and HCCA legacy}
%   \subsection{TXOP Reservation and Trigger-based MU-EDCA}

% % 3 ─────────── Real-time theory
% \section{Real-Time Scheduling Theory}
%   \subsection{Periodic, Sporadic and Aperiodic Servers (Deferrable, Sporadic, Polling)}
%   \subsection{Fixed-Priority vs. EDF and Utilisation Bounds}

% % 4 ─────────── TSN toolbox
% \section{Time-Sensitive Networking (TSN)}
%   \subsection{IEEE 802.1AS / PTP Clock Synchronisation}
%   \subsection{IEEE 802.1Qbv Time-Aware Shaper (TAS)}
%   \subsection{Credit-Based Shaper and Cyclic Queuing \& Forwarding}

% % 6 ─────────── V2X constraints
% \section{V2X Safety-Service Requirements}
%   \subsection{Latency, Reliability and Vehicle Density Targets}
%   \subsection{URLLC Benchmarks and Comparison with C-V2X}


% \section{IEEE\,802.11p Foundations for V2X QoS}
% \label{sec:rel_11p}

% IEEE 802.11p—branded as Dedicated Short-Range Communications (DSRC)—was created to let vehicles broadcast time-critical safety information such as Basic Safety Messages (BSM) and Cooperative Awareness Messages (CAM).  These packets must reach nearby cars and roadside units within a few tens of milliseconds even when no infrastructure exists.  To meet that goal, 802.11p modifies legacy 802.11a in four decisive ways\cite{80211p}:

% \begin{enumerate}[label=\arabic*)]
% \item \textbf{5.9 GHz band with 10 MHz channels.}  Halving the channel width doubles the OFDM symbol duration, which in turn improves resilience to multipath and Doppler at highway speeds.
% \item \textbf{Longer guard interval (1.6 µs).}  The extended cyclic prefix handles delay spreads up to \(\approx\!300\) ns, common in roadside reflections.
% \item \textbf{Outside-the-Context-of-a-BSS (OCB) or “WAVE mode.”}  Frames carry a wildcard BSSID, so stations exchange data without association, authentication, or beacon overhead—crucial when contact time is measured in seconds.
% \item \textbf{Prioritised EDCA access.}  Safety messages map to the highest access categories, pre-empting infotainment traffic during contention.
% \end{enumerate}

% Time is slotted into fixed 50 ms Synchronisation Intervals: all nodes first listen on the Control Channel for safety beacons, then hop to a Service Channel for non-safety exchanges.  Combined, these changes keep end-to-end latency under 100 ms and sustain \(\gtrsim\!90\%\) packet-delivery ratios at several-hundred-metre ranges, establishing the baseline against which newer Wi-Fi 6/7 and 5G-V2X QoS schemes are judged.



% % 8 ─────────── Evaluation platform
% \section{Simulation Framework}
%   \subsection{OMNeT++/INET Wi-Fi Models (ac/ax)}
%   \subsection{Channel, Mobility and PHY/MAC Limitations}

% \section{Bounded Latency in IEEE~802.11}
% \label{sec:bounded_latency}

% 


\section{End-to-End Delay in IEEE 802.11 DCF}
\label{sec:background_dcf_delay}

To understand the feasibility of bounded-latency communication over IEEE~802.11 networks, we examine the analytical delay framework presented in \emph{On the End-to-End Delay Analysis of the IEEE 802.11 Distributed Coordination Function}~\cite{vardakas2007end}, which builds upon Bianchi's saturation model for the Distributed Coordination Function (DCF). Their work decomposes the total end-to-end delay into one equation that has two distinct components:

\begin{equation}
D = Q + M
\end{equation}

where:
\begin{itemize}
    \item $Q$: \textit{Queuing delay} - Time spent waiting in the transmission buffer
    \item $M$: \textit{MAC delay} - Time from head-of-queue to successful transmission
\end{itemize}

\subsection{Key Components of Delay}

\noindent\textbf{MAC Delay ($M$)} is dominated by:
\begin{itemize}
    \item \textit{Backoff process}: Random waiting periods using binary exponential backoff
    \item \textit{Collision handling}: Retransmissions with increasing contention windows
    \item \textit{Channel contention}: Frozen backoff counters during others' transmissions
\end{itemize}

\noindent\textbf{Queuing Delay ($Q$)} follows the M/G/1 model:
\begin{equation}
\E[Q] = \frac{\rho \E[M]}{2(1-\rho)} \left(1 + \frac{(\E[M])^2}{\Var[M]}\right)
\end{equation}
where:
\begin{itemize}
    \item $\rho$: Traffic load intensity (utilization)
    \item $\E[M]$: Mean MAC delay
    \item $\Var[M]$: MAC delay variance
    \item The term $\left(1 + \frac{(\E[M])^2}{\Var[M]}\right)$ is the \textit{form factor of the holding time distribution}.
\end{itemize}

\subsection{Critical System Parameters}

\begin{itemize}
    \item \textit{Number of stations ($n$)}: Delay grows super-linearly with $n$ due to increased collisions
    \item \textit{Access mechanism}:
    \begin{itemize}
        \item \textit{Basic access}: Efficient at high data rates
        \item \textit{RTS/CTS}: Reduces collisions at low rates but adds control overhead
    \end{itemize}
    \item \textit{Data rate}: Higher rates reduce transmission time but control frames (ACK/RTS/CTS) remain fixed at base rate
    \item \textit{Contention parameters}: $CW_{\min}$, $CW_{\max}$, and retry limit $m$
\end{itemize}

\subsection{Delay Characteristics}

The MAC delay distribution exhibits:
\begin{itemize}
    \item Heavy-tailed characteristics due to exponential backoff
    \item High variance under congestion ($\Var[M]$ increases with collision probability)
    \item Sensitivity to initial contention window settings
\end{itemize}

\subsection{Practical Implications}
\begin{itemize}
    \item RTS/CTS beneficial only at low data rates (1 Mbps)
    \item Basic access preferable at higher rates (5.5/11 Mbps)
    \item Small networks favor smaller $CW_{\min}$ for lower latency
    \item Large networks require larger $CW_{\min}$ to reduce collisions
    \item Control frame overhead becomes bottleneck at high data rates
    \item Minimizing $\Var[M]$ is crucial for predictable latency at high loads
\end{itemize}

The fundamental tension in DCF latency stems from its distributed nature: While avoiding centralized control, the random access mechanism introduces inherent latency variability that grows with network size and load.